biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
papertra knit and minor updates

---
classoption: "runningheads"
title: "Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
shorttitle: "Estimating the SE impacts of car substitution by bicycle and public transit"

@@

*biclar* builds on the Propensity to Cycle Tool (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^2]
However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.

[^2]:  See <https://www.npt.scot> and <https://cruse.bike> for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.

@@

The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types, and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring].
`r5r` can calculate multi-modal routes using PT combined with other modes.
It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^3] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.

[^3]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)

@@

The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data agregated and validated.
This information is crucial for an accurate PT trip and route estimation.
A digital elevation model from the European Space Agency's COPERNICUS mission, with a 25m spatial resolution, was used to include street gradient information, as a weight in cycling routing.
The cycling potential trips for the two ENMAC targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.

@@

Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
using the HEAT for Cycling and the HEAT as a Service[^4] tools, from the WHO.

[^4]: see https://github.com/HEAT-WHO/HEAT_heatr_api

@@

By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas.
The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.

@@

Please place your acknowledgments at the end of the paper, preceded by an unnumbered run-in heading (i.e. 3rd-level heading).

>>>>

---
classoption: "runningheads"
title: "biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
shorttitle: "biclaR: SE impacts of car substitution by bicycle and public transit"

@@

*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.

[^2]: See [pct.bike](https://www.pct.bike/)

[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.

@@

The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types, and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring].
`r5r` can calculate multi-modal routes using PT combined with other modes.
It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.

[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)

@@

The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data agregated and validated.
This information is crucial for an accurate PT trip and route estimation.
A digital elevation model from the European Space Agency's COPERNICUS mission, with a 25m spatial resolution, was used to include street gradient information, as a weight in cycling routing.
The cycling potential trips for the two national strategic targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.

@@

Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
using the HEAT for Cycling and the HEAT as a Service[^5] tools, from the WHO.

[^5]: see https://github.com/HEAT-WHO/HEAT_heatr_api

@@

By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas.

The train with most potential (as seen from fig 2)

The information available at the open access website can be downloaded, and used with any GIS software to, for instance, understant which cycling connections have the highest socioenvironmental impacts, in tons of avoided CO2eq emissions, or in long term social costs.

The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.

@@

### 2023-09-27 paper/PaperTRA/PaperTRA.tex
papertra knit and minor updates

\title{Estimating the socio-environmental impacts of car substitution by
bicycle and public transit using open tools}
%
\titlerunning{Estimating the SE impacts of car substitution by bicycle
and public transit}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Rosa Félix\inst{1}\orcidID{0000-0002-5642-6006} \and Filipe
Moura\inst{1}\orcidID{0000-0001-7749-8490} \and Robin
Lovelace\inst{2}\orcidID{0000-0001-5679-6536}}

@@

\emph{biclar} builds on the Propensity to Cycle Tool (PCT), a web
application and research project funded by the UK's Department for
Transport in 2015 which launched nationally in 2017 as part of the
government's Cycling and Walking Investment Strategy. The PCT initially
used only origin-destination data for commuting trips as the basis of
estimates of cycling potential at zone, route and route network levels
{[}\textbf{Lovelace2017?}{]}.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}4{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}5{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}6{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data agregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model
from the European Space Agency's COPERNICUS mission, with a 25m spatial
resolution, was used to include street gradient information, as a weight
in cycling routing. The cycling potential trips for the two ENMAC
targets (4\% and 10\%) was estimated from the values for cycling and car
trips (both as a driver and as a passenger) from the 2017 base scenario.

@@

{[}Tabela 11{]}

@@

{[}Tabela 13{]}

@@

Please place your acknowledgments at the end of the paper, preceded by
an unnumbered run-in heading (i.e.~3rd-level heading).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}

>>>>

\title{biclaR: Estimating the socio-environmental impacts of car
substitution by bicycle and public transit using open tools}
%
\titlerunning{biclaR: SE impacts of car substitution by bicycle and
public transit}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Rosa Félix\inst{1}\orcidID{0000-0002-5642-6006} \and Filipe
Moura\inst{1}\orcidID{0000-0001-7749-8490} \and Robin
Lovelace\inst{2}\orcidID{0000-0001-5679-6536}}

@@

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}4{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}5{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}6{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data agregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model
from the European Space Agency's COPERNICUS mission, with a 25m spatial
resolution, was used to include street gradient information, as a weight
in cycling routing. The cycling potential trips for the two national
strategic targets (4\% and 10\%) was estimated from the values for
cycling and car trips (both as a driver and as a passenger) from the
2017 base scenario.

@@

\[Tabela 11\]

@@

\[Tabela 13\]

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
to school in england, and associated health and carbon impacts:
Application of the {`}propensity to cycle tool{'}. Journal of Transport
\& Health. 12, 263--278 (2019).
https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}
